\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-18 paper.tex
Added model interpretations for MNIST digit 1 data set

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, boxrule=0pt, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

@@

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training PBMC3k dataset. What is the effective number of
bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
80, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\begin{figure}[h]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP-1.pdf}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?, (b) Model in the 2D
space with UMAP (\url{https://youtu.be/CNXjWZMyh_M})}

@@

\begin{figure}[h]

@@

\caption{\label{fig-nldervisPBMCUMAP3}(a) 2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for the
PBMC3k dataset. Is this a best representation of the original data?, (b)
Model in the 2D space with UMAP (\textless\textgreater)}

@@

\begin{itemize}
\tightlist
\item
  Digit 1 (7877 images only)
\end{itemize}

@@

\begin{itemize}
\item
  Total number of bins used: 180 (num\_x = 12, num\_y = 15, hex\_size =
  0.06)
\item
  Learned from the model: There is a non-linear shape in 2D, but if look
  at the model in high-D, the non-linear continuity is there but there
  is a twist in the structure which didn't capture by PaCMAP.
\item
  Write about looking at images of digit 1 as well..how in 2D space
  images are distributed, which images occur high-errors in high-D
\end{itemize}

@@

\begin{verbatim}
$mse
[1] 2.276023

$aic
[1] 66382.78
\end{verbatim}

\begin{figure}[h]

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a
best representation of the original data?, (b) Model in the 2D space
with PaCMAP ()}

@@

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{story}{%
\subsection{Story}\label{story}}

\begin{itemize}
\tightlist
\item
  UMAP representation of PBMC3k data with 15 PCs and parameter choice of
  n:neighbor:30, min\_dist:0.1, metric:cosine (author's suggestion)
  shows three well-separated clusters. The question we need to answer
  is: Does this UMAP representation preserve the actual data structure,
  or is something missing or misidentified or misleading?
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMPview-1.pdf}

\caption{\label{fig-nldervisPBMCUMPview}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?}

\begin{itemize}
\item
  To investigate this, we are going to visualise the model in the data
  space, which means that the model generated from UMAP is visualised in
  high-D with original data.
\item
  To fit the model, first we need to decide how many number bins should
  allocate along the x and y axes. To make an initial choice, we use MSE
  plot.
\end{itemize}

@@

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training PBMC3k dataset. What is the effective number of
bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
588, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\begin{itemize}
\tightlist
\item
  Then, the model is constructed by following all the steps in 2D and
  lifts the model into high-D.
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP-1.pdf}

@@

\caption{\label{fig-nldervisPBMCUMAP}Model in the 2D space with UMAP
(\url{https://youtu.be/CNXjWZMyh_M})}

\begin{itemize}
\item
  You will see three well-separated clusters in high-D as well which
  gives evidence that UMAP preserves the global structure. Furthermore,
  there is non-linearity within the clusters which was not identified by
  the UMAP, which proves that local structure is not preserved.
\item
  Can we find a better parameter choice for UMAP which preserve the
  local structure as well?
\item
  To do that, we are going to fit models for different parameter choice
  and select the one which have the lowest error.
\end{itemize}

\begin{itemize}
\item
  Selected the one with n:neighbor:5, min\_dist:0.99, metric:cosine.
\item
  Now, we are going to fit the model for the new parameter choice.
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAPview2-1.pdf}

\caption{\label{fig-nldervisPBMCUMAPview2}(a) 2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for the
PBMC3k dataset. Is this a best representation of the original data?}

@@

@@

\caption{\label{fig-nldervisPBMCUMAP3}Model in the 2D space with UMAP
(\textless\textgreater)}

@@

\begin{itemize}
\item
  Why I choose this dataset? Because I need show that whatever NLDR
  technique that we use, our model can capture the non-linear structure
  within clusters as well.
\item
  Digit 1 (7877 images only)
\end{itemize}

@@

\begin{itemize}
\item
  Total number of bins used: 180 (num\_x = 12, num\_y = 15, hex\_size =
  0.06)
\item
  Learned from the model: There is a non-linear shape in 2D, but if look
  at the model in high-D, the non-linear continuity is there but there
  is a twist in the structure which didn't capture by PaCMAP in 2D
  layout.
\end{itemize}

\hypertarget{story-1}{%
\subsubsection{Story}\label{story-1}}

\begin{itemize}
\item
  PaCMAP representation of MNIST digit 1 data with 10PCs and parameter
  choice of n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0 shows a non-linear structure (see
  Figure~\ref{fig-nldervisMNISTPACMAP} (a)). I selected this parameter
  choice because this the default parameter setting for PaCMAP and
  selected PaCMAP because when I observed the 5 NLDR techniques applied
  for the data PaCMAP showed the non-linear structure effectively in 2D.
\item
  By looking at MSE plot, I selected the 180 bins, which associated with
  12 bins along the x-axis, 15 bins along the y-axis, and hex size as
  0.06 (Figure~\ref{fig-diagnosticpltMNIST} (b)). Then, fitted the model
  in 2D (see Figure~\ref{fig-nldervisMNISTPACMAP} (b)). As you can see
  in Figure~\ref{fig-nldervisMNISTPACMAP} (b), the fitted model capture
  the non-linear structure of the 2D data.
\end{itemize}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-84-1.pdf}

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from PaCMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset, (b) Model
in the 2D space overlaid on PaCMAP data, and (c) Error occurred by the
high-D model for high-D data and the error mapped into 2D NLDR points.}

@@

\begin{itemize}
\tightlist
\item
  Next, the constructed 2D model lifts into high-dimensions (see
  Figure~\ref{fig-mnist1_sc}).
\end{itemize}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_1.png}

\subcaption{\label{fig-mnist1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_2.png}

\subcaption{\label{fig-mnist1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_3.png}

\subcaption{\label{fig-mnist1_sc3}}
\end{minipage}%

\caption{\label{fig-mnist1_sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model in high-D, a video of the
tour animation is available at (\url{https://youtu.be/zq2GM9qvUNA}).}

\begin{itemize}
\item
  As shown in Figure~\ref{fig-mnist1_sc} (a), the non-linear continuity
  structure that you have seen in 2D representation of PaCMAP, also can
  be seen in the model in high-D space as well. This gives us an
  evidence that our model capture the non-linear structure of the data
  effectively.
\item
  Furthermore, our model shows some addition pattern of data which
  couldn't be seen in the 2D layout.
\item
  One is the twisted structure within the non-linear structure (see
  Figure~\ref{fig-mnist1_sc} (c)).
\item
  Another thing is the connected corners of the non-linear structure
  (see Figure~\ref{fig-mnist1_sc} (b)). The non-linear curve is bended.
\item
  To gives the evidence that PaCMAP preserve the local structure
  (non-linear pattern), our model helps by computing the model errors
  (Figure~\ref{fig-diagnosticpltMNIST} (c)).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  As you can see in Figure~\ref{fig-diagnosticpltMNIST} (c), the largest
  model error shows in the upper right side of the 2D layout. A sample
  of the images that shows the largest error can be seen in
  Figure~\ref{fig-sampleimg}. The reason for the large model error is
  that the high-D points related to this largest error located away from
  the high-D model points.
\end{itemize}
